Unifying intra- and inter-specific variation in tropical tree mortality

### 2017-05-31 ms/ms.Rnw
forgot to save these small changes
Edits throughout
- refine and shorten text
- update PNAS style
- better formatting of Supp Matt
- update units in figures

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

@@

\title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{James S Camac}
\author[a]{Mark Westoby}
\author[b]{S Joseph Wright}
\author[a]{Daniel S Falster}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}

@@

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{Tree mortality is a fundamental demographic process which affects forest dynamics and consequently, a terrestrial ecosystems ability to store carbon. Here, for the first time, we use observed survival records over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent (i.e. death by stocastic chance) mortality across 203 tropical forest tree species. We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island (BCI). Furthermore, we found that wood density most strongly influenced mortality rates by decreasing growth-independent mortality. However, wood density only accounted for a small fraction of the overall species variability observed in growth-independent mortality rates, suggesting that there must be other species traits that strongly affect mortality. Which traits and how many of them, however, remains unknown.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.C. designed and performed the analysis; D.F. conceived the idea, supervized the development of this work; M.W. provided biological insight. All authors contributed to the writing of this manuscript.}

\authordeclaration{The authors declare that they have no competing financial interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: james.camac@gmail.com}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits and temporal effects -- these distinct drivers are yet to be integrated within a cohesive and predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop mortality functions that not only integrate individual, species and temporal effects, but also simultaneously estimate the relative contribution of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate. We show that functions that partition mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability observed between species, and that temporal variation in mortality was large. Lastly, our analysis indicate that growth-dependent processes are the predominant contributor to rates of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests. More importantly, it highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

\begin{document}
% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalized the article contents.
\verticaladjustment{-2pt}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.

@@

Two difficulties arise when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates\cite{Poorter:2008iu}. This requirement arises in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis\cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$\cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required. As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in monitoring plots containining thousands of individuals\cite{Anderson-Teixeira-2015a}.

The other difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}. Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}. Yet, such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}. This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.

\section*{A unified model of tree mortality}

A particular challenge in developing hazard functions for plants is to integrate the conflicting relationships observed between mortality and growth within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies broadly support this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. By contrast, across species, there tends to be a strong trade-off between growth and survival, with faster growing species typically exhibiting higher mortality rates than slower growing species\cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events\cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

Another major challenge is estimating the relative contribution of growth-dependent and growth-independent hazards on an individual's overall instantaneous mortality rate. While plants can be killed by many factors, they all broadly fall into two independent hazards. Plants can die as a result of insufficient carbon assimulation required for growth and repair (i.e. growth-dependent hazards), or they can die because of stochastic events that are independent of an individual's growth rate (e.g.  windfall, fire). By developing models that can simultaneously estimate growth-dependent and growth-independent hazards we can begin to examine their relative contribution and how these vary by ecosystem. More importantly, such models will allow us to determine how plant traits and other factors influence each hazard. For example, wood density is thought to primarily affect growth-independent mortality by reducing stem breakage \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}. However, recent research has also indicated that higher wood density may convey increased tolerance to shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}. The question is, does wood density affect both hazards? And if so, which hazard does it affect most? Answering such questions will ultimately allow us to incorporate more biological information into existing mortality functions, which in turn, will increase model predictive capacity.

\begin{figure}[!hb]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, alive or dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
B) We consider three alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard; and 3) a function that combines both baseline and growth-dependent hazards. The parameters of the models are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote, or baseline hazard. Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g. windfall, fire) that kill a plant irrespective of its growth rate. For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
C) Log-loss function highlighting that the further the predicted probability is from truth the higher the logloss.
D) The predictive accuracy of alternative models is evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data. We then average the estimated log loss for each of these 10 model fits.}
\label{fig1}
\end{figure}

In this study, we reconcile both intraspecific and interspecific factors and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards. We achieve this by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework. To derive this unified model, we compare the capacity of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases. Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality rate decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth (i.e. growth-dependent hazard) and death due to stochastic events such as windthrow or fire (i.e. growth-independent or baseline hazard). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}. However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census and growth-dependent and growth-independent hazards?
  \item Does mortality rates vary substantially between censuses?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used 10-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over a two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).

@@

\subsection*{Mortality over time}
Tree mortality progressively decreased over time. Between 1995 and 2010, the highest proportion of trees death occured between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive accuracy (i.e. lower logarithmic loss; Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, then our model predicted deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

\begin{figure}[!h]
\includegraphics{figures/fig2.pdf}
\caption{Mean ($\pm$ 95\% credible intervals) logarithmic loss whereby lower values equal higher predictive accuracy. 
A) Five sequential stages of model selection with each stage increasing in complexity. Null: null model ($\gamma$); Census: inclusion of census effects ($\delta$); Growth rate: the inclusion of growth rate. This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{\beta X_{i}})\delta$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{\beta X_{i}} + \gamma)\delta$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
B) Change in logloss under all wood density\textendash parameter combinations. Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth. Symbols represent general functional form: null/baseline (circle), growth\textendash dependent hazard model (triangle) and model incorporating both baseline and growth\textendash dependent hazards (square). Note: errors are too small to see in panel A.}
\label{fig2}
\end{figure}

\subsection*{Growth hazard functions}
We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e. $(\alpha e^{\beta X_{i}} + \gamma)\delta$), significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e. $(\alpha e^{\beta X_{i}})\delta$) hazard functions (Fig. \ref{fig2}A). Moreover, we found that predictive accuracy was higher for stem diameter growth relative to basal-area growth. Perhaps surprisingly, we detected no significant improvement in predictive accuracy between the baseline only and growth-dependent only hazard functions.

@@

\subsection*{Effect of wood density on mortality rates}
The inclusion of wood density on either baseline or growth-dependent parameters significantly improved model accuracy relative to a hazard function without such effects (Fig. \ref{fig2} A,B). However, the most parsimonious model with highest predictive accuracy (i.e. lowest logarithmic loss) was that which attributed the wood density effect, $\rho$, to the baseline hazard term ($\gamma$; Fig. \ref{fig2} B). Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate (Fig. \ref{fig3}). This meant that fast growing individuals from a low wood density species had on higher average mortality rates ((Fig. \ref{fig4} A,B), and thus higher annual probability of death (Fig. \ref{fig4} C,D), relative to fast growing individuals of high wood density species.

@@

\begin{figure}[h!]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in instantaneous mortality rates (A,B), and its derivative, annual mortality probability (C,D), with increasing growth rates. A,C) Estimated species curves for all 203 BCI species used in this study. B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.}
\label{fig4}
\end{figure}

\subsection*{Mortality rates, maximum species size and species light requirements}
We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were positively correlated with gap index (our measure of species light requirement; Fig. \ref{fig6} A,C, E). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. The combination of these correlations suggest that species which predominately recruit in gaps (i.e. higher gap index) were more prone to death due to both low growth and stochastic chance. They also required faster growth rates to achieve a given mortality rate relative to a species that recruits in non-gaps. We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig. \ref{fig6} B,D,F,H).

@@

The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards\cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}. We show that regardless of growth measure, incorporating both hazards significantly improves model predictive accuracy. This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality. The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stressors (e.g. competition, parasitization, localized herbivory). By contrast, the growth-independent hazard (i.e. baseline hazard) accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g. windthrow).

@@

Lastly, research should also examine how different growth measures influence predictive accuracy of mortality models. We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.

>>>>

\documentclass[9pt,twocolumn,twoside]{pnas-new}\usepackage[]{graphicx}\usepackage[]{color}

% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

@@

\title{Unifying intra- and inter-specific variation in tropical tree mortality}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{James S Camac}
\author[a]{Mark Westoby}
\author[b]{S Joseph Wright}
\author[a,c]{Daniel S Falster}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
\affil[c]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales}

@@

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling. Here, for the first time, we use observed survival records over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent (i.e. stochastic) mortality across 203 tropical forest tree species. We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island. Furthermore, we found that wood density most strongly influenced mortality rates by decreasing growth-independent mortality. However, wood density only accounted for a small fraction of the overall species variability observed in growth-independent mortality rates, suggesting that there must be other species traits that strongly affect mortality. Which traits and how many of them, however, remains unknown.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.C. performed research; D.S.F. conceived the idea; S.J.W. collected data. All authors contributed to the writing of this manuscript.}

\authordeclaration{The authors declare that they have no competing financial interests.\\

Data deposition: Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
}

\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: james.camac@gmail.com}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales, these distinct drivers are yet to be integrated within a unified, predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop mortality functions that not only integrate individual, species and temporal effects, but also simultaneously estimate the relative contribution of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate. We show that functions that partition mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability observed between species, and that temporal variation in mortality was large. Lastly, our analysis indicate that growth-dependent processes are the predominant contributor to rates of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests. More importantly, it highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalized the article contents.
\verticaladjustment{-2pt}

@@

@@

Two difficulties arise when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates\cite{Poorter:2008iu}. This requirement arises in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$, also called the ``hazard function''. In classic survival analysis\cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_2$ and $t_3$ is a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$\cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are long-lived and we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required. As such, detailed studies of tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals\cite{Anderson-Teixeira-2015a}.

Another difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of hazard functions that skilfully predict patterns in mortality. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}. Generally, however, these influences have not been integrated into a common hazard function\cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.

\section*{Towards a unified model of tree mortality}

A specific challenge in developing hazard functions for plants is to estimate the relative contribution of growth-dependent and growth-independent hazards on an individual's overall hazard. While plants die via many causes, these broadly fall into two categories. The first are growth-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair. The second are growth-independent hazards, where plants die because of stochastic events, irrespective of their growth rate, such as  windfall or fire. To capture this distinction, we assume an individual's hazard is the sum of growth-independent ($\lambda_{\textrm{b},i}(t)$) and growth-dependent ($\lambda_{\textrm{g},i}(t)$) rates:
\begin{equation} \label{eq:additive}
\lambda_i(t)  = \lambda_{\textrm{b},i}(t) + \lambda_{\textrm{g},i}(t).
\end{equation}

A further challenge in developing hazard functions for plants is to integrate the conflicting relationships observed within and across species between mortality and growth. Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies broadly support this theory, with many indicating exponential declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, such that with parameters $\alpha, \beta$:
\begin{equation} \label{eq:gd}
\lambda_{\textrm{g},i}(t) = \alpha \textrm{e}^{-\beta X_i(t)}.
\end{equation}

By contrast, across species, there is a strong trade-off between growth and survival, with individuals from faster-growing species  exhibiting higher mortality rates than individuals from slower-growing species\cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density. Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

\begin{figure}[!hb]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, alive or dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
B) We consider three alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard; and 3) a function that combines both baseline and growth-dependent hazards. The parameters of the models are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote, or baseline hazard. Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g. windfall, fire) that kill a plant irrespective of its growth rate. For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
C) Log-loss function highlighting that the further the predicted probability is from truth the higher the logloss.
D) The predictive skill of alternative models is evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data. We then average the estimated log loss for each of these 10 model fits.}
\label{fig1}
\end{figure}

In this study, we attempt to reconcile intra-specific and inter-specific factors as well as partition instantaneous mortality rates into growth-dependent and growth-independent rates. We achieve this by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework. To derive this unified model, we compare the skill of three alternative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. Beginning from eq.\ref{eq:additive}, the simplest model (Fig. \ref{fig1}B, left), assumes a single baseline, growth-independent hazard rate. The second model (Fig. \ref{fig1}B, middle) assumes the risk of dying eventually declines to zero as growth increases (eq. \ref{eq:gd}). The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models, which allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\geq 0$.  Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.

To derive the unified model, we asked the following questions:
1) Does mortality rates vary substantially between censuses, such that a specific census effect is needed?
2) Does a hazard function including both growth-dependent and independent terms outperform a function including only one of these?
3) Which growth measure (stem diameter or stem area increment) more skilfully predicts growth-dependent mortality?
4) Does including wood density (a species trait) improve model skill, and if so, does it affect parameters of growth-dependent or independent hazard or both?
5) What proportion of inter-specific variability in hazard is explained by wood density?
6) What proportion of total variability in hazard is explained by species, census, growth-dependent and growth-independent hazards?

Answering these questions involves comparing the  predictive skill of alternative models. To achieve this, we used 10-fold cross-validation (Fig. \ref{fig1}D), to systematically assesses each model's skill in predicting outcomes in novel data (i.e. not used in model calibration) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across five iterative stages of model development, corresponding to the six questions evaluated above, of gradually increased complexity (see Methods).

@@

\subsection*{Mortality over time}
Comparing the 3 intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive skill (i.e. Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

\begin{figure}[!h]
\includegraphics{figures/fig2.pdf}
\caption{Mean ($\pm$ 95\% credible intervals) logarithmic loss whereby lower values equal higher predictive skill.
A) Five sequential stages of model selection with each stage increasing in complexity. Null: null model ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate. This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
B) Change in logloss under all wood density\textendash parameter combinations. Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth. Symbols represent general functional form: null/baseline (circle), growth\textendash dependent hazard model (triangle) and model incorporating both baseline and growth\textendash dependent hazards (square). Note: errors are too small to see in panel A.}
\label{fig2}
\end{figure}

\subsection*{Growth hazard functions}
We found that a hazard function with both growth-dependent and growth-independent terms (i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), significantly outperformed both the baseline only (i.e.$\gamma \delta_t$) or growth-dependent only (i.e. $(\alpha e^{-\beta X_{i}})\delta_t$) functions (Fig. \ref{fig2}A). Moreover, we found that predictive skill was higher when using stem diameter growth over basal-area growth.

@@

\subsection*{Effect of wood density on mortality rates}
Including wood density ($\rho$) as an effect on either baseline or growth-dependent parameters significantly improved model skill relative to a hazard function without such effects (Fig. \ref{fig2} A,B). The most parsimonious model, with highest predictive skill, was that which attributed the wood density effect to the baseline hazard term ($\gamma$; Fig. \ref{fig2} B). Wood density was therefore negatively correlated with a species' baseline mortality rate (Fig. \ref{fig3}).

% This meant that fast growing individuals from a low wood density species had on higher average mortality rates (Fig. \ref{fig4} A,B), and thus higher annual probability of death (Fig. \ref{fig4} C,D), relative to fast growing individuals of high wood density species.

@@

\subsection*{Mortality rates, maximum species size and species light requirements}
We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were weakly and positively correlated with gap index (our measure of species light requirement; Fig. \ref{fig6} A,C, E). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. The combination of these correlations suggest that species which predominately recruit in gaps (i.e. higher gap index) were more prone to death due to both low growth and stochastic chance. They also required faster growth rates to achieve a given mortality rate relative to a species that recruits in non-gaps. We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig. \ref{fig6} B,D,F,H).

@@

The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards\cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}. We show that regardless of growth measure, incorporating both hazards significantly improves model predictive skill. This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality. The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stressors (e.g. competition, parasitization, localized herbivory). By contrast, the growth-independent hazard (i.e. baseline hazard) accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g. windthrow).

@@

Lastly, research should also examine how different growth measures influence predictive skill of mortality models. We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive skill, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
